---
title: "2017_centralcity_microbiome_analyses"
author: "Yen Ning Chai"
date: "2023-09-05"
output:
  html_document:
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library("phyloseq")
library("ggplot2")
library("plyr")
library("scales")
library(grid)
library(reshape2)
library(ape)
library(vegan)
library(emmeans)
library(ggpubr)
library(rstatix)
library(multcompView)
library(xtable)
library(PMCMRplus)
library(tidyverse)
```

# Fig 2

## Fig2A: Unconstrained PCoA for all sample types/compartments (soil between rows, soil within rows, rhizosphere, root, and leaf)

```{r fig2a}
ss.map <- read.delim("map_cc_0718_samples_but_blanks.txt") #change this
row.names(ss.map)<-ss.map$SampleID
ss.map$SampleID<- NULL
ss.map$GrowthCondtion <- factor(ss.map$GrowthCondtion)
ss.map$Plate <- as.factor(ss.map$Plate)
ss.map$Rep <- as.factor(ss.map$Rep)
ss.map$Plot <- as.factor(ss.map$Plot)
ss.map$Variety <- as.factor(ss.map$Variety)
ss.map$SampleType <- as.factor(ss.map$SampleType)

ss.bcurtis <- read.table("distance-matrix.tsv", header = T, row.names = 1)
ss.bcurtis <- ss.bcurtis[match(row.names(ss.map), row.names(ss.bcurtis)), match(row.names(ss.map), colnames(ss.bcurtis))]

#PCoA
r <- pcoa(ss.bcurtis)
r.axes <- data.frame((scores(r$vectors[,1:2])))
comp.col <- c("blue","purple","red","orange","darkgreen","yellow","purple","grey","green")
plot<-ggplot(r.axes, aes(x = Axis.1, y = Axis.2))+ #label=rownames(bcurtis.cap.whole.axes)+ #change, or can add "size = Replicate"
  geom_vline(xintercept = 0, alpha = 0.9) +
  geom_hline(yintercept = 0, alpha = 0.9) +
  geom_point(size = 3, alpha = 0.75, aes(color=ss.map$SampleType, shape=ss.map$GrowthCondtion)) +
  theme_classic() +
  labs(x = "PCo1 (15.95%)", y = "PCo1 (8.56%)") + #change
  scale_color_manual(values = comp.col) +
  theme(text = element_text(size = 20)) +
  #geom_text(size=3)+
  #stat_ellipse()+
  labs(color = "Sample Type", shape="Treatment") +
  #guides(fill = guide_legend(title = "Growth Condition")) +
  theme(panel.border = element_rect(colour = "black",fill=NA))

adonis(ss.bcurtis~ SampleType*GrowthCondtion*Variety,data = ss.map,add = T, strata = ss.map$Rep)

ggsave(filename="pcoa_sampletype_trt.png", plot=plot, device="png", height=5, width=7.5, units="in", dpi=500)
```

## Fig 2B: Alpha diversities

```{r fig2b}
## faith's pd
alpha <- read.delim("faith_pd_cc.txt") 
alpha$SampleType <- factor(alpha$SampleType, levels = c("Soil Between Rows","Soil Within Row","Rhizosphere","Root","Leaf"))
alpha$SampleType_GrowthCondtion <- factor(alpha$SampleType_GrowthCondtion, levels=c("Soil Between Rows_Full N","Soil Between Rows_Low N","Soil Within Row_Full N","Soil Within Row_Low N","Rhizosphere_Full N","Rhizosphere_Low N","Root_Full N","Root_Low N","Leaf_Full N","Leaf_Low N"))

a<-aov(faith_pd ~ SampleType_GrowthCondtion, data = alpha)
tukey<-TukeyHSD(a)

kruskal.test(faith_pd ~ SampleType_GrowthCondtion, data = alpha)
pw<-pairwise.wilcox.test(alpha$faith_pd, alpha$SampleType_GrowthCondtion, p.adjust.method = "fdr")

pw
out.p <- get.pvalues(pw)
out.mcV <- multcompLetters(out.p,compare="<",threshold=0.05,Letters=letters)
out.mcV


bxp <- ggboxplot(alpha, x = "SampleType_GrowthCondtion", y = "faith_pd", ylim=range(0,250), fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), notch = TRUE)

ggsave(filename="faith_pd.png", plot=bxp, device="png", height=5, width=6, units="in", dpi=500)

## shannon
alpha <- read.delim("shannon_cc.txt")
alpha$SampleType <- factor(alpha$SampleType, levels = c("Soil Between Rows","Soil Within Row","Rhizosphere","Root","Leaf"))
alpha$SampleType_GrowthCondtion <- factor(alpha$SampleType_GrowthCondtion, levels=c("Soil Between Rows_Full N","Soil Between Rows_Low N","Soil Within Row_Full N","Soil Within Row_Low N","Rhizosphere_Full N","Rhizosphere_Low N","Root_Full N","Root_Low N","Leaf_Full N","Leaf_Low N"))

kruskal.test(shannon_entropy ~ SampleType_GrowthCondtion, data = alpha)
pw<-pairwise.wilcox.test(alpha$shannon_entropy, alpha$SampleType_GrowthCondtion, p.adjust.method = "fdr")
pw
out.p <- get.pvalues(pw)
out.mcV <- multcompLetters(out.p,compare="<",threshold=0.05,Letters=letters)
out.mcV


bxp <- ggboxplot(alpha, x = "SampleType_GrowthCondtion", y = "shannon_entropy", ylim=range(0,15), fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), notch = TRUE)

ggsave(filename="shannon.png", plot=bxp, device="png", height=5, width=6, units="in", dpi=500)

```


## Fig 2C: Relative abundance

```{r fig3c}
ps = import_biom("ASV_table_all_sample_without_blank__Date_71817_with_tax.biom", treefilename="rooted-tree_cc.nwk", refseqfilename="ASVs_sequence.fa")
mapfile <- import_qiime_sample_data("map_cc_0718_samples_but_blanks__Date_71817__.txt")

mydata <- merge_phyloseq(ps, mapfile)
ps.rarefied = rarefy_even_depth(mydata, rngseed=2, sample.size=10240, replace=F)

physeq2 = filter_taxa(ps.rarefied, function(x) mean(x) > 0.1, TRUE)
physeq3 = transform_sample_counts(physeq2, function(x) x / sum(x) )
physeq3

glom <- tax_glom(physeq3, taxrank = 'Rank2')
glom.trt <- merge_samples(glom, "SampleType_GrowthCondtion")
glom.trt.phylum = transform_sample_counts(glom.trt, function(x) x / sum(x))
data_glom <- psmelt(glom.trt.phylum) # create dataframe from phyloseq object

trt <- ggplot(data=data_glom, aes(x=SampleType_GrowthCondtion, y=Abundance, fill=Rank2)) +
facet_grid(~SampleType, scales = "free") +
geom_bar(aes(),stat="identity", position="stack") +
scale_fill_manual(values = c("darkblue", "darkgoldenrod1", "darkseagreen", "mediumpurple4", "darkolivegreen1", "lightskyblue", "darkgreen", "indianred1", "black", "red4", "aquamarine", "darkorange1", "cyan1", "cornflowerblue", "chocolate4","red2","cyan4","bisque3","darkolivegreen2","deeppink1","honeydew3","khaki","lightblue4","lightsalmon3","magenta1","lightblue","gold1","purple1","royalblue4","thistle","yellow","lavender","grey")) +
theme(legend.position="bottom")

ggsave(filename="relativeabundance_sampletype_trt.png", plot=trt, device="png", height=6, width=10, units="in", dpi=1000)
```

# Fig 3
## Fig A-E: Constrained analysis of principal coordinates in each sample type/compartment (soil between rows, soil within rows, rhizosphere, root, and leaf)

```{r fig 3}
ss.map <- read.delim("map_cc__Date_71817____SampleType_Soil__.txt") #change this
row.names(ss.map)<-ss.map$SampleID
ss.map$SampleID<- NULL
ss.map$GrowthCondtion <- factor(ss.map$GrowthCondtion)
ss.map$Plate <- as.factor(ss.map$Plate)
ss.map$Rep <- as.factor(ss.map$Rep)
ss.map$Plot <- as.factor(ss.map$Plot)


ss.bcurtis <- read.table("distance-matrix_full.tsv", header = T, row.names = 1)
ss.bcurtis <- ss.bcurtis[match(row.names(ss.map), row.names(ss.bcurtis)), match(row.names(ss.map), colnames(ss.bcurtis))]


bcurtis.cap.whole <- capscale(as.dist(ss.bcurtis) ~ Variety*GrowthCondtion+Condition(Rep),data = ss.map, add = T)
bcurtis.cap.whole.axes <- data.frame(cbind(ss.map, scores(bcurtis.cap.whole)$sites))
#wuf.cap.whole.axes$Plant <- factor(wuf.cap.whole.axes$Plant,levels = c("Adenum","Aloe","Cissus","Cleome","Zygophyllum"))#change
#comp.col <- c("red","blue","black","darkgreen","darkgreen","blue","purple","orange","yellow","purple","grey","green")
comp.col <- c("red2","cyan4","bisque3","darkolivegreen2","deeppink1","honeydew3","khaki","lightblue4","lightsalmon3","magenta1","lightblue","gold1","purple1","royalblue4","thistle")

percent_explained <- bcurtis.cap.whole$CCA$eig / sum(bcurtis.cap.whole$CCA$eig) * 100
plot<-ggplot(bcurtis.cap.whole.axes, aes(x = CAP1, y = CAP2, color=Variety, shape=GrowthCondtion) )+ #change, or can add "size = Replicate"
  geom_vline(xintercept = 0, alpha = 0.9) +
  geom_hline(yintercept = 0, alpha = 0.9) +
  geom_point(size = 4, alpha = 0.75) +
  theme_classic() +
  labs(x = "Constrained PCo1 (56.60%)", y = "Constrained PCo2 (17.08%)") + #change
  scale_color_manual(values = comp.col) +
  #scale_shape_manual(values = comp.shape) +
  labs(color="Sorghum Type", shape="Treatment") +
  theme(text = element_text(size = 20)) +
  theme(panel.border = element_rect(colour = "black",fill=NA))


adonis2(ss.bcurtis~ Variety+GrowthCondtion,data = ss.map,add = T, strata = ss.map$Rep)
```

# Fig 4

```{r}
d <- read.csv("ra_sampletype_trt.csv", header=TRUE, sep=",") 

d$SampleType_GrowthCondtion <- as.factor(d$SampleType_GrowthCondtion)
d$SampleType <- as.factor(d$SampleType)

d$SampleType_GrowthCondtion <- factor(d$SampleType_GrowthCondtion, levels=c("Soil Between Rows_Full N","Soil Between Rows_Low N","Soil Within Row_Full N","Soil Within Row_Low N","Rhizosphere_Full N","Rhizosphere_Low N","Root_Full N","Root_Low N","Leaf_Full N","Leaf_Low N"))

d$SampleType <- factor(d$SampleType, levels=c("Soil Between Rows","Soil Within Row","Rhizosphere","Root","Leaf"))

bxp1 <- ggboxplot(d, x = "SampleType_GrowthCondtion", y = "Bacteroidota", fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), notch=TRUE)



bxp1<- bxp1 + ggtitle("Bacteroidota") + 
  geom_bracket(xmin = "Rhizosphere_Full N", xmax = "Rhizosphere_Low N", y.position = 0.3,label = "**") + 
  geom_bracket(xmin = "Root_Full N", xmax = "Root_Low N", y.position = 0.7,label = "*") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
  theme(panel.border = element_rect(colour = "black",fill=NA), text = element_text(size=15)) +
  theme(axis.text=element_text(size=11)) +
  rotate_x_text(60) + labs(x = "Sample Type_Treatment", y = "Relative Abundance") 


bxp2 <- ggboxplot(d, x = "SampleType_GrowthCondtion", y = "Firmicutes", fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), ylim=c(0,0.5), notch=TRUE)


bxp2<- bxp2 + labs(title = "Firmicutes") +
  geom_bracket(xmin = "Root_Full N", xmax = "Root_Low N", y.position = 0.17,label = "**") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
  theme(panel.border = element_rect(colour = "black",fill=NA), text = element_text(size=15)) +
  theme(axis.text=element_text(size=11)) +
  rotate_x_text(60) + labs(x = "Sample Type_Treatment", y = "Relative Abundance") 

  
bxp3 <- ggboxplot(d, x = "SampleType_GrowthCondtion", y = "Myxococcota", fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), ylim=c(0,0.25), notch=TRUE)


bxp3<- bxp3 + labs(title = "Myxococcota") +
  geom_bracket(xmin = "Rhizosphere_Full N", xmax = "Rhizosphere_Low N", y.position = 0.1,label = "*",tip.length = 0.008) + 
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
  theme(panel.border = element_rect(colour = "black",fill=NA), text = element_text(size=15)) +
  theme(axis.text=element_text(size=11)) +
  rotate_x_text(60) + labs(x = "Sample Type_Treatment", y = "Relative Abundance") 


bxp4 <- ggboxplot(d, x = "SampleType_GrowthCondtion", y = "Acidobacteriota", fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), ylim=c(0,0.25), notch=TRUE)


bxp4<- bxp4 + labs(title = "Acidobacteriota") +
  geom_bracket(xmin = "Root_Full N", xmax = "Root_Low N", y.position = 0.17,label = "**",tip.length = 0.01) +
  theme(panel.border = element_rect(colour = "black",fill=NA), text = element_text(size=15)) +
  theme(axis.text=element_text(size=11)) +
  rotate_x_text(60) + labs(x = "Sample Type_Treatment", y = "Relative Abundance")


bxp5 <- ggboxplot(d, x = "SampleType_GrowthCondtion", y = "Chloroflexi", fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), ylim=c(0,0.15), notch=TRUE)


bxp5<- bxp5 + labs(title = "Chloroflexi") +
  geom_bracket(xmin = "Rhizosphere_Full N", xmax = "Rhizosphere_Low N", y.position = 0.1,label = "*",tip.length = 0.01) + 
  geom_bracket(xmin = "Root_Full N", xmax = "Root_Low N", y.position = 0.1,label = "**",tip.length = 0.01) +
  theme(panel.border = element_rect(colour = "black",fill=NA), text = element_text(size=15)) +
  theme(axis.text=element_text(size=11)) +
  rotate_x_text(60) + labs(x = "Sample Type_Treatment", y = "Relative Abundance") 


bxp6 <- ggboxplot(d, x = "SampleType_GrowthCondtion", y = "Crenarchaeota", fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), ylim=c(0,0.2), notch=TRUE)


bxp6<- bxp6 + labs(title = "Crenarchaeota") +
  geom_bracket(xmin = "Root_Full N", xmax = "Root_Low N", y.position = 0.05,label = "*") +
  theme(panel.border = element_rect(colour = "black",fill=NA), text = element_text(size=15)) +
  theme(axis.text=element_text(size=11)) +
  rotate_x_text(60) + labs(x = "Sample Type_Treatment", y = "Relative Abundance") 

bxp7 <- ggboxplot(d, x = "SampleType_GrowthCondtion", y = "Planctomycetota", fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), ylim=c(0,0.1), notch=TRUE)



bxp7<- bxp7 + labs(title = "Planctomycetota") +
  geom_bracket(xmin = "Rhizosphere_Full N", xmax = "Rhizosphere_Low N", y.position = 0.06,label = "*") + 
  geom_bracket(xmin = "Root_Full N", xmax = "Root_Low N", y.position = 0.07,label = "***") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
  theme(panel.border = element_rect(colour = "black",fill=NA), text = element_text(size=15)) +
  theme(axis.text=element_text(size=11)) +
  rotate_x_text(60) + labs(x = "Sample Type_Treatment", y = "Relative Abundance") 


bxp8 <- ggboxplot(d, x = "SampleType_GrowthCondtion", y = "Gemmatimonadota", fill = "SampleType", palette =c("blue","purple","red","orange","darkgreen"), ylim=c(0,0.06), notch=TRUE)


bxp8<- bxp8 + labs(title = "Gemmatimonadota") +
  geom_bracket(xmin = "Root_Full N", xmax = "Root_Low N", y.position = 0.038,label = "*") +
  theme(panel.border = element_rect(colour = "black",fill=NA), text = element_text(size=15)) +
  theme(axis.text=element_text(size=11)) +
  rotate_x_text(60) + labs(x = "Sample Type_Treatment", y = "Relative Abundance") 

```

# Fig 5

## The relationship between bacterial alpha diversities in rhizosphere and sorghum NUE.

```{r fig5}
# shannon
## all types
data <- read.csv("shannon_biomasss_ratio_mean_correlation_root.csv",header=TRUE, sep=",")
shannon <- data$shannon_ratio
biomass <- data$biomass_ratio

cor <- cor.test(shannon, biomass, method = c("kendall"))
summary(cor)

plot<-ggplot(data,aes(shannon, biomass_ratio)) +
#geom_point(size = 3, alpha = 0.75) +
geom_point(size = 3, alpha = 0.75,aes(colour = factor(Type))) +
scale_color_manual(values = c("Energy" = "red2", "Grain" = "cyan4","Sweet" = "bisque3")) +
geom_smooth(method=lm, color="black") + 
labs(x = "Shannon Diversity Ratio", y = "Dry Biomass Ratio") +
annotate("text", x = 1.15, y = 1.1, label =expression(italic('τ')*' = 0.10, '*italic('P')*' = 0.53')) +
labs(color="Sorghum Type") +
theme_classic() + 
theme(panel.border = element_rect(colour = "black",fill=NA),panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
theme(text = element_text(size = 20)) + theme(axis.text.y = element_text(color="black", size=15),axis.text.x = element_text(color="black", size=15))

ggsave(filename="cor_biomass_shannon_ratio_root.png", plot=plot, device="png", height=5, width=7.5, units="in", dpi=500)

#energy and grain only
data <- read.csv("shannon_biomasss_ratio_mean_correlation_root_excluded_sweet.csv",header=TRUE, sep=",")
shannon <- data$shannon_ratio
biomass <- data$biomass_ratio

cor <- cor.test(shannon, biomass, method = c("kendall"))
summary(cor)

plot<-ggplot(data,aes(shannon, biomass_ratio)) +
#geom_point(size = 3, alpha = 0.75) +
geom_point(size = 3, alpha = 0.75,aes(colour = factor(Type))) +
scale_color_manual(values = c("Energy" = "red2", "Grain" = "cyan4")) +
geom_smooth(method=lm, color="black") + 
labs(x = "Shannon Diversity Ratio", y = "Dry Biomass Ratio") +
annotate("text", x = 1.15, y = 1.1, label =expression(italic('τ')*' = 0.10, '*italic('P')*' = 0.53')) +
labs(color="Sorghum Type") +
theme_classic() + 
theme(panel.border = element_rect(colour = "black",fill=NA),panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
theme(text = element_text(size = 20)) + theme(axis.text.y = element_text(color="black", size=15),axis.text.x = element_text(color="black", size=15))

ggsave(filename="cor_biomass_shannon_ratio_root_excluded_sweet.png", plot=plot, device="png", height=5, width=7.5, units="in", dpi=500)

# faith's pd
# all types
data <- read.csv("faithpd_biomasss_ratio_mean_correlation_root.csv",header=TRUE, sep=",")
faithpd <- data$faith_ratio
biomass <- data$biomass_ratio

cor <- cor.test(faithpd, biomass, method = c("kendall"))
summary(cor)

plot<-ggplot(data,aes(faithpd, biomass_ratio)) +
#geom_point(size = 3, alpha = 0.75) +
geom_point(size = 3, alpha = 0.75,aes(colour = factor(Type))) +
scale_color_manual(values = c("Energy" = "red2", "Grain" = "cyan4","Sweet" = "bisque3")) +
geom_smooth(method=lm, color="black") + 
labs(x = "Faith's PD Ratio", y = "Dry Biomass Ratio") +
annotate("text", x = 1.2, y = 1.1, label =expression(italic('τ')*' = 0.19, '*italic('P')*' = 0.23')) +
labs(color="Sorghum Type") +
theme_classic() + 
theme(panel.border = element_rect(colour = "black",fill=NA),panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
theme(text = element_text(size = 20)) + theme(axis.text.y = element_text(color="black", size=15),axis.text.x = element_text(color="black", size=15))

ggsave(filename="cor_biomass_faithpd_ratio_root.png", plot=plot, device="png", height=5, width=7.5, units="in", dpi=500)

#energy and grain only
data <- read.csv("faithpd_biomasss_ratio_mean_correlation_root_excluded_sweet.csv",header=TRUE, sep=",")
faithpd <- data$faith_ratio
biomass <- data$biomass_ratio

cor <- cor.test(faithpd, biomass, method = c("kendall"))
summary(cor)

plot<-ggplot(data,aes(faithpd, biomass_ratio)) +
#geom_point(size = 3, alpha = 0.75) +
geom_point(size = 3, alpha = 0.75,aes(colour = factor(Type))) +
scale_color_manual(values = c("Energy" = "red2", "Grain" = "cyan4")) +
geom_smooth(method=lm, color="black") + 
labs(x = "Faith's PD Ratio", y = "Dry Biomass Ratio") +
annotate("text", x = 1.21, y = 1.1, label =expression(italic('τ')*' = 0.10, '*italic('P')*' = 0.53')) +
labs(color="Sorghum Type") +
theme_classic() + 
theme(panel.border = element_rect(colour = "black",fill=NA),panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
theme(text = element_text(size = 20)) + theme(axis.text.y = element_text(color="black", size=15),axis.text.x = element_text(color="black", size=15))

ggsave(filename="cor_biomass_faithpd_ratio_root_excluded_sweet.png", plot=plot, device="png", height=5, width=7.5, units="in", dpi=500)

```

# Fig 6

## Bacterial taxa affected by sorghum NUE.

```{r}
data <- read.csv("pseudomonas_mean_biomass_ra_ratio_correlation.csv",header=TRUE, sep=",")
cor <- lm(biomass_ratio~Pseudomonas_ratio, data=data)
ra <- data$Pseudomonas_ratio
biomass <- data$biomass_ratio

cor <- cor.test(ra, biomass, method = c("kendall"))

plot<-ggplot(data,aes(Pseudomonas_ratio, biomass_ratio)) +
geom_point(size = 3, alpha = 0.75) +
#geom_point(size = 3, alpha = 0.75,aes(colour = factor(Type))) +
#scale_color_manual(values = c("Energy" = "black", "Grain" = "red","Sweet" = "blue")) +
geom_smooth(method=lm, color="black") + 
labs(x = expression(italic('Pseudomonas ')*'Fold Change'), y = "Dry Biomass Ratio") +
annotate("text", size = 5, x = 1.7, y = 0.9, label =expression(italic('τ')*' = -0.47, '*italic('P')*' = 0.008')) +
theme_classic() + 
theme(panel.border = element_rect(colour = "black",fill=NA),panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
theme(text = element_text(size = 20)) + theme(axis.text.y = element_text(color="black", size=15),axis.text.x = element_text(color="black", size=15))

ggsave(filename="cor_biomass_pseudomonas_ra_mean_ratio.png", plot=plot, device="png", height=5, width=5, units="in", dpi=500)

```


```{r}

```

