---
title: "2017 Sorghum Data - Biomass"
author: "Emily Goren"
date: "12/2/2017"
output:
  html_document:
    highlight: tango
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = c("png", "pdf"), cache = TRUE)
options(scipen = 999)
require(ggplot2)
require(lmerTest)
require(emmeans)
require(pbkrtest)
require(knitr)
theme_set(theme_bw(base_size = 16) + theme(legend.position = "bottom"))
```

# Data

```{r data}
# Function to extract line type from genotype.
get.line <- function(genotypes) sapply(genotypes, function(geno) {
  switch(substr(geno, 1, 1), "E" = "Energy", "S" = "Sweet", "G" = "Grain")
})
# Central city data.
cc <- read.csv("cc_biomass_extract_EG.csv")
cc$dry <- cc$Total.Dry.Weight.kg.acre
cc$fresh <- cc$Total.Fresh.Weight.kg.acre
cc$block <- as.factor(cc$Rep)
cc$line <- get.line(cc$Geno)
str(cc)

# Subset by location and line.
cc.E <- subset(cc, line == "Energy")
cc.S <- subset(cc, line == "Sweet")
cc.G <- subset(cc, line == "Grain")

```


### Biomass histograms

```{r histograms, fig.width=11}
ggplot(cc, aes(x = fresh)) +
  geom_histogram(bins = 25) +
  facet_grid(Treat ~ line, scales = "fixed") +
  labs(x = "Total Fresh Weight (kg/acre)") +
  ggtitle("Central City Fresh Weight Histograms")
ggplot(cc, aes(x = dry)) +
  geom_histogram(bins = 25) +
  facet_grid(Treat ~ line, scales = "fixed") +
  labs(x = "Total Dry Weight (kg/acre)") +
  ggtitle("Central City Dry Weight Histograms")
```


# Model

Analyze as a split-plot design with blocking on "replicate." If no evidence of a block effect, remove block. Note that genotypes were not completely randomized within treatment, as some genotypes were randomized to the southmost or northmost plots within a block based on height.


### Fit linear mixed model (random block effects) 

- For each combination of weight type, line, and location assume that $Y_{ijk} = \mu_{ij} + \tau_{k} + \gamma_{ik} + \varepsilon_{ijk}$, where 
    - $Y_{ijk}$ is the (possibly log transformed) response for treatment $i = 1,2$; genotype $j = 1, \dots, J$; block $k = 1, \dots, 8$.
    - $\mu_{ij} = \alpha_i + \beta_j + (\alpha\beta)_{ij}$ are fixed effects describing the genotype-by-treatment mean.
    - $\tau_k \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2_\tau)$ are block random effects.
    - $\gamma_{ik} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2_\gamma)$ are block-by-treatment random effects.
    - $\varepsilon_{ijk} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2_\varepsilon)$ are random errors.

- Test for block effect ($H_0: \sigma^2_\tau = 0$ versus $H_1: \sigma^2_\tau > 0$).

```{r lmm}
# CC fresh
cc.fresh.E <- lmer(log(fresh) ~ Treat*Geno + (1|Treat:block) + (1|block), data = cc.E)
cc.fresh.S <- lmer(log(fresh) ~ Treat*Geno + (1|Treat:block) + (1|block), data = cc.S)
cc.fresh.G <- lmer(log(fresh) ~ Treat*Geno + (1|Treat:block) + (1|block), data = cc.G)
# CC dry
cc.dry.E <- lmer(log(dry) ~ Treat*Geno + (1|Treat:block) + (1|block), data = cc.E)
cc.dry.S <- lmer(log(dry) ~ Treat*Geno + (1|Treat:block) + (1|block), data = cc.S)
cc.dry.G <- lmer(log(dry) ~ Treat*Geno + (1|Treat:block) + (1|block), data = cc.G)

# Summarize.
lmm <- list(cc.fresh.E, cc.fresh.G, cc.fresh.S,
            cc.dry.E, cc.dry.G, cc.dry.S)
l <- c("Central City - Fresh - Energy", 
       "Central City - Fresh - Grain", 
       "Central City - Fresh - Sweet",
       "Central City - Dry - Energy", 
       "Central City - Dry - Grain", 
       "Central City - Dry - Sweet")
trans <- c(rep("Log", 6))
vars <- data.frame(Analysis = l,
                   Transformation = trans,
                   BlockVariance = sapply(lmm, function(f) as.numeric(VarCorr(f)$block)),
                   #p.value = sapply(lmm, function(f) rand(f)$rand.table["block", "p.value"]/2),
                   TrtBlkVariance = sapply(lmm, function(f) as.numeric(VarCorr(f)$'Treat:block')),
                   ErrorVariance = sapply(lmm, function(f) attr(VarCorr(f), "sc")^2))
kable(vars, digits = 4)

```


### Remove block effects where appropriate

- No evidence of a block effect ($p-val > 0.1$) for 
    - Central City - Fresh - Grain
    - Central City - Dry - Energy
    - Central City - Dry - Grain

- For the above analyses, remove random block effect and fit the reduced linear mixed model $Y_{ijk} = \mu_{ij} + \gamma_{ik} + \varepsilon_{ijk}$.

```{r lm}
# CC fresh
cc.fresh.G <- lmer(log(fresh) ~ Treat*Geno + (1|Treat:block), data = cc.G)
# CC dry
cc.dry.E <- lmer(log(dry) ~ Treat*Geno + (1|Treat:block), data = cc.E)
cc.dry.G <- lmer(log(dry) ~ Treat*Geno + (1|Treat:block), data = cc.G)

```


### Check residual plots

```{r resids, fig.width = 10}
fit <- list(cc.fresh.E, cc.fresh.G, cc.fresh.S,
            cc.dry.E, cc.dry.G, cc.dry.S)
n <- unlist(lapply(fit, function(f) length(fitted(f))))
# Function to extract treatment from fitted model object.
get.trt <- function(obj) {
  if (class(obj) == "lm") out <- obj$model$Treat else out <- obj@frame$Treat
  return(out)
}
resids <- data.frame(Analysis = rep(l, n),
                     Transformation = rep(trans, n),
                     Fitted = unlist(lapply(fit, fitted)),
                     Residual = unlist(lapply(fit, resid)),
                     Treatment = unlist(lapply(fit, get.trt)))
resids$Response <- ifelse(sapply(resids$Analysis, function(f) grepl("Dry", f)), "Dry Weight", "Fresh Weight")
resids$Location <- ifelse(sapply(resids$Analysis, function(f) grepl("Central City", f)), "Central City")
resids$Line <- sapply(as.character(resids$Analysis), function(f) substr(f, nchar(f)-5, nchar(f)))
# Plot.
ggplot(subset(resids, Location == "Central City" & Response == "Fresh Weight"), 
       aes(y = Residual, x = Fitted, color = Treatment)) +
  ggtitle("Central City - Fresh Weight") +
  geom_point() + scale_color_grey(start = 0.6, end = 0) +
  facet_wrap( ~ Line, scales = "free")
ggplot(subset(resids, Location == "Central City" & Response == "Dry Weight"), 
       aes(y = Residual, x = Fitted, color = Treatment)) +
  ggtitle("Central City - Dry Weight") +
  geom_point() + scale_color_grey(start = 0.6, end = 0) +
  facet_wrap( ~ Line, scales = "free")
```

# Estimates

Find the ratio of estimated means (low/high) for each genotype with 95\% confidence intervals (on original, not log, scale). See output at end or exported .csv file or all estimates and $p$-value for the test of a treatment effect (i.e., whether the ratio is equal to one).

For analysis performed on the original scale (no log transformation applied) as done for the energy lines at Scottsbluff, the confidence interval was computed using a generalization of Fieller 1954 (same methodology used for the 2016 data). 

If blocks were included in the model, degrees of freedom were computed using the Kenward-Roger method implemented in the emmeans package.


```{r ratios}
# Function to find confidence bounds using generalization of Fieller 1954.
find.T.bd <- function(theta, a1, a2, beta.hat, var.beta.hat, alpha, df, bd = 'lower') {
    c <- a1 - theta * a2
    L <- t(c) %*% beta.hat
    var.L <- t(c) %*% var.beta.hat %*% c
    T <- drop(L / sqrt(var.L))
    crit <-  qt(1-alpha/2, df)
    if (bd == 'lower') {
        out <- crit - T
    } else {
        out <- - crit - T
    }
    return(out)
}
# Function to get CI for ratio of high/low treatment means and p-value for H0: ratio = 1.
# Uses Kenward-Roger method to estimate df if block is included in model.
get.res <- function(fit) {
  mixed <- class(fit) != "lm"
  if (mixed) log <- grepl("log", fit@call[2]) else log <- grepl("log", fit$call[2])
  if (log) {
    if(mixed) b <- fixef(fit) else b <- coef(fit)
    m <- ifelse("TreatWW" %in% names(b), "pairwise", "revpairwise")
    high <- ifelse("TreatWW" %in% names(b), "WW", "Full N")
    low <- ifelse("TreatWW" %in% names(b), "WS", "Low N")
    lsm <- emmeans(fit, ~ Treat | Geno)
    ests <- confint(contrast(lsm, method = m))
    emm <- summary(lsm)
    out <- data.frame(Genotype = ests$Geno)
    out$Est.High.log <- emm$emmean[emm$Treat == high]
    out$SE.High.log <- emm$SE[emm$Treat == high]
    out$Est.High <- exp(out$Est.High.log)
    out$SE.High <- NA
    out$Est.Low.log <- emm$emmean[emm$Treat == low]
    out$SE.Low.log <- emm$SE[emm$Treat == low]
    out$Est.Low <- exp(out$Est.Low.log)
    out$SE.Low <- NA
    out$Treatment <- gsub(" - ", "/", ests$contrast)
    out$Ratio <- exp(ests$estimate)
    out$Lower.95.CL <- exp(ests$lower.CL)
    out$Upper.95.CL <- exp(ests$upper.CL)
    out$df <- ests$df
    out$p.value <- summary(contrast(lsm, method = m))$p.value
    out$Log <- log
  } else {
    lsm <- emmeans(fit, ~ Treat | Geno)
    if(mixed) b <- fixef(fit) else b <- coef(fit)
    var.b <- as.matrix(vcov(fit))
    A <- lsm@linfct
    effects <- colnames(A)
    if(mixed) genolev <- fit@frame$Geno else genolev <- fit$model$Geno
    genos <- paste0("Geno", unique(genolev))
    tmp <- lapply(genos, function(g) {
      if (g %in% effects) { # Genotype isn't reference level.
        contrs <- A[A[,g] == 1,]
        if ("TreatWW" %in% effects) {
          a1 <- contrs[contrs[,"TreatWW"] == 0,]
          a2 <- contrs[contrs[,"TreatWW"] == 1,]
        } else {
          a2 <- contrs[contrs[,"TreatLow N"] == 0,]
          a1 <- contrs[contrs[,"TreatLow N"] == 1,]
        }
      } else { # Genotype is reference level.
        nonrefs <- genos[genos != g]
        contrs <- A[apply(A[,nonrefs], 1, function(f) all(f == 0)),]
        if ("TreatWW" %in% effects) {
          a1 <- contrs[contrs[,"TreatWW"] == 0,]
          a2 <- contrs[contrs[,"TreatWW"] == 1,]
        } else {
          a2 <- contrs[contrs[,"TreatLow N"] == 0,]
          a1 <- contrs[contrs[,"TreatLow N"] == 1,]
        }
      }
      low <- a1 %*% b
      low.se <- sqrt(t(a1) %*% var.b %*% a1)
      high <- a2 %*% b
      high.se <- sqrt(t(a2) %*% var.b %*% a2)
      val <- low / high
      diffs <- summary(contrast(lsm, method = "pairwise"))
      idx <- diffs$Geno == gsub("Geno", "", g)
      DF <- diffs$df[idx]
      r1 <- uniroot(find.T.bd, maxiter = 1e6,
                    lower = -1e6, upper = 1e6,
                    a1 = a1, a2 = a2,
                    beta.hat = b, var.beta.hat = var.b,
                    alpha = 0.05, df = DF, bd = 'lower')
      r2 <- uniroot(find.T.bd, maxiter = 1e6,
                    lower = -1e6, upper = 1e6,
                    a1 = a1, a2 = a2,
                    beta.hat = b, var.beta.hat = var.b,
                    alpha = 0.05, df = DF, bd = 'upper')
      ans <- data.frame(Genotype = gsub("Geno", "", g),
                        Est.High.log = NA,
                        SE.High.log = NA,
                        Est.High = high,
                        SE.High = high.se,
                        Est.Low.log = NA,
                        SE.Low.log = NA,
                        Est.Low = low,
                        SE.Low = low.se,
                        Treatment = ifelse("TreatWW" %in% effects, "WS/WW", "Low N/Full N"),
                        Ratio = val,
                        Lower.95.CL = r1$root,
                        Upper.95.CL = r2$root,
                        df = DF,
                        p.value = diffs$p.value[idx])
      return(ans)
    })
    out <- do.call("rbind", tmp)
    out$Log <- log
  }
  num <- as.numeric(unlist(regmatches(out$Genotype, gregexpr("[[:digit:]]+", out$Genotype))))
  return(out[order(num),])
}
results <- lapply(fit, get.res)
names(results) <- l
```


### Plots of ratio of estimated mean biomass (low/high treatment) with 95% confidence intervals


```{r plotratios, fig.width = 10}
results2 <- results
for (r in 1:length(results2)) {
  results2[[r]]$Line <- get.line(results2[[r]]$Genotype)
  results2[[r]]$Response <- ifelse(grepl("Dry", names(results2)[r]), "Dry Weight", "Fresh Weight")
  results2[[r]]$Location <- ifelse(grepl("Central City", names(results2)[r]), "Central City")
}
results2 <- do.call("rbind", results2)
write.csv(results2, "2017_sorghum_biomass_estimates_splitplot_noE10atSB.csv", row.names = FALSE)
ggplot(subset(results2, Location == "Central City" & Response == "Fresh Weight"),
       aes(x = Genotype, y = Ratio, fill = Line)) + 
  geom_bar(stat = "identity") + scale_fill_brewer(palette = "Dark2") +
  geom_errorbar(aes(ymin = Lower.95.CL, ymax = Upper.95.CL), width = 0.5) + # 95% CI
  geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +
  xlab("Genotype") + ylab("Ratio of Mean Fresh Weight \n (Low/High N)") +
  ggtitle("Central City - Fresh Weight")
ggplot(subset(results2, Location == "Central City" & Response == "Dry Weight"),
       aes(x = Genotype, y = Ratio, fill = Line)) + 
  geom_bar(stat = "identity") + scale_fill_brewer(palette = "Dark2") +
  geom_errorbar(aes(ymin = Lower.95.CL, ymax = Upper.95.CL), width = 0.5) + # 95% CI
  geom_hline(yintercept = 1, color = "darkgrey", linetype = "dashed") +
  xlab("Genotype") + ylab("Ratio of Mean Dry Weight \n (Low/High N)") +
  ggtitle("Central City - Dry Weight")

```



### Plots of estimated mean weight (kg/acre) for each genotype and treatment combination

```{r plotmeans}
results3 <- reshape(results2, direction = "long", 
                    varying = c("Est.Low", "Est.High"))
iscc <- results3$Location == "Central City"
results3$Treatment[iscc] <- ifelse(results3$time[iscc] == "High", "Full N", "Low N")

# CC fresh
ggplot(subset(results3, 
              Location == "Central City" & Response == "Fresh Weight" & Line == "Energy"),
       aes(x = Genotype, y = Est, fill = Treatment)) + 
  geom_bar(stat = "identity", position = position_dodge()) + 
  scale_fill_grey(start = 0.6, end = 0) +
  xlab("Genotype") + ylab("Estimated Mean Fresh Weight (kg/acre)") +
  ggtitle("Central City - Fresh Weight - Energy") 
ggplot(subset(results3, 
              Location == "Central City" & Response == "Fresh Weight" & Line == "Grain"),
       aes(x = Genotype, y = Est, fill = Treatment)) + 
  geom_bar(stat = "identity", position = position_dodge()) + 
  scale_fill_grey(start = 0.6, end = 0) +
  xlab("Genotype") + ylab("Estimated Mean Fresh Weight (kg/acre)") +
  ggtitle("Central City - Fresh Weight - Grain") 
ggplot(subset(results3, 
              Location == "Central City" & Response == "Fresh Weight" & Line == "Sweet"),
       aes(x = Genotype, y = Est, fill = Treatment)) + 
  geom_bar(stat = "identity", position = position_dodge()) + 
  scale_fill_grey(start = 0.6, end = 0) +
  xlab("Genotype") + ylab("Estimated Mean Fresh Weight (kg/acre)") +
  ggtitle("Central City - Fresh Weight - Sweet") 
# CC dry
ggplot(subset(results3, 
              Location == "Central City" & Response == "Dry Weight" & Line == "Energy"),
       aes(x = Genotype, y = Est, fill = Treatment)) + 
  geom_bar(stat = "identity", position = position_dodge()) + 
  scale_fill_grey(start = 0.6, end = 0) +
  xlab("Genotype") + ylab("Estimated Mean Dry Weight (kg/acre)") +
  ggtitle("Central City - Dry Weight - Energy") 
ggplot(subset(results3, 
              Location == "Central City" & Response == "Dry Weight" & Line == "Grain"),
       aes(x = Genotype, y = Est, fill = Treatment)) + 
  geom_bar(stat = "identity", position = position_dodge()) + 
  scale_fill_grey(start = 0.6, end = 0) +
  xlab("Genotype") + ylab("Estimated Mean Dry Weight (kg/acre)") +
  ggtitle("Central City - Dry Weight - Grain") 
ggplot(subset(results3, 
              Location == "Central City" & Response == "Dry Weight" & Line == "Sweet"),
       aes(x = Genotype, y = Est, fill = Treatment)) + 
  geom_bar(stat = "identity", position = position_dodge()) + 
  scale_fill_grey(start = 0.6, end = 0) +
  xlab("Genotype") + ylab("Estimated Mean Dry Weight (kg/acre)") +
  ggtitle("Central City - Dry Weight - Sweet") 

```


### Big tables of estimates 

Exported as .csv file "2017_sorghum_biomass_estimates_splitplot_noE10atSB.csv.csv"

```{r printratios, results = "asis"}
for (r in 1:length(results)) {
  print(kable(results[[r]], digits = 3, row.names = FALSE,
              caption = names(results)[r]), padding = 0)
}
```